---
title: "JSC370 Midterm"
output: html_document
date: "2024-03-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

## Introduction

\
In recent years there has been an increasing pessimism about the prospect of 
owning a home, with younger generations dissatisfied how older generations
own homes well they cannot afford to. Younger generations may question whether
they will ever be able to own their own homes, and compare themselves to 
the seemingly far favorable conditions for older generations; indeed, home 
ownership and housing affordability is often cited as a core issue for
younger generations [3].

\
The goal of this project is to 
examine differences in home ownership between generations through home ownership rates by 
year, age, and locations. The proposed research question is: Have home ownership rates in 
the United States changed between the generations of Baby Boomers, Gen X, Millenials, 
and Gen Z?

\
The dataset that was used to answer this question is a subset of the 
IPSUM CPS dataset. The dataset comes from the United States
Census Current Population Survey (CPS) and its supplemental Annual Social and 
Economic Supplement survey (ASEC). The CPS is preformed monthly to different
households in the US while the ASEC supplement is preformed annualy in March.
The dataset contains both household and individual level information, meaning
some variables refer to entire household information such as whether a house
is owned or rented, while other variables refer to individual information
such as a persons age.

\
As the entire IPUMS CPS dataset is large and contains many variables
that are irrelevant to this analysis, a subset of the data was used containing
only the variables and year range of interest described in the methods section.


## Methods

The methods section includes subsections corresponding to how the data was
obtained, data cleaning, data wrangling, and the methods for how the plots
were produced.

### Data Collection

\
IPUMS CPS allows users to create
their own subsets, called data extracts, of the full CPS dataset that contain
variables and year ranges of interest. To create a data extract, a user must
select variables and year ranges and then wait for a data extract to
be generated; once the extract is generated it can be downloaded. The creation
and retrieval of the data extract used in this analysis was preformed through
the IPUMS API. To create the extract through the API, a post request was made
using the httr R package with a formatted JSON string that specified the year range
and variables of interest. The year range was 1976-2023 and the variables of 
interest were:

\
-YEAR which is the year an observation was from

-AGE which is the age of an individual,

-OWNERSHIP which is whether the household that an individual belongs to is rented/owned, 

-RELATE which is the relation to their relation to the householder of the household an individual belongs to

-STATEFIP which is the FIP code of the state were the observation was from

-COUNTY which is the FIP code of the county were the observation was from

\
Once the data extract request was submitted, a get 
request was used to check whether the data extract had been generated. Finally,
when the data extract was complete, the data files were downloaded from the
URLs provided from the content of the get request. The dataset was then loaded
in R using the ipumsr package provided by IPUMS for loading IPUMS data into
R.

\
In addition to the key variables mentioned in the introduction, IPUMS provides
weights for calculating individual and household level statistics to mitigate
the sampling bias of the the survey, these weights were also included in the
dataset as the variables ASECWTH for household level statistics
and ASECWT for individual level statistics.

### Data Cleaning

\
Once the dataset was loaded into R, it's dimensions were checked for how
many observations were present in the dataset and whether the number of variables
was consistent with what was requested in the data extract. Checking the dimension
showed that the dataset contains 8,293,750 observations with each observation being an 
individual who belongs to a household. The head, tail, and variable types were also 
checked to make sure the dataset was imported correctly.

\
Once the dataset was confirmed to have been imported correctly, the variables
were checked for missing values and suspicious values. All values for AGE,
YEAR, RELATE, OWNERSHP, and STATEFIP had no missing values and were all in
their expected range; however, the variable COUNTY had 3,159,092 missing values
and 2,991,865 values of 0 which is not a recognized county FIP code. Checking
the IPUMS data dictionary confirmed that a value of 0 was used for counties that
did not meet a population threshold due to concerns about deanonymizing
data. Further investigation also showed that all the missing values in the COUNTY
variable occurred before 1995. Removing observations with missing values for
COUNTY would remove a significant amount of the data. Additionally, the COUNTY
variable was not used across most of the analysis; therefore, no observations
were removed.

\
Something to note in the dataset is that the values in the categorical variables
are given by codes that correspond to their meaning, for example, the variable
OWNERSHIP has a value of "10" when the observation belongs to a household that 
is owned. The meaning of these codes is easy to find on the IPUMS CPS website,
so these values were not changed in this analysis.

### Data Wrangling

\
The number of observations with age >90 were small which created outliers in
the home ownership rates within these age ranges. Additionally, In the United States, 
people cannot own property until they reach the age of majority which is 19 in some states. 
Therefore, the age ranges of <19 and >90 were removed from the data set. After removing
all observations with age less than 19 and greater than 90 there were 5,796,909 observations left.
Removing these entries means that the population of interest in this analysis are
adults in the United States between the ages of 19 and 90.

\
For the analysis, a new variable had to be created referring the generation that individual belongs to. 
There are no official age cutoffs for what defines a generation, so instead
the generation age ranges were chosen to be consistent with an 
article by Pew Research Center, the age ranges were: 1946-1964 (Baby Boomer),
1965-1980 (Gen X), 1981-1996 (Millennial), 1997-2012 (Gen Z) [2]. The generation
variable was created by subtracting the AGE variable from the YEAR variable to
get a birth year and then checking which generation range it was in. Finally,
the frequencies of each generation was checked to see if the variable was
create properly, as you might expect, Gen Z had the least amount of observations 
at 86,029 while Baby Boomers had the most 2,147,536	due to the data set containing
many years before any individual in Gen Z was born. 


### Methods for Analysis

\
Once the data cleaning and wrangling was complete, the analysis could be preformed. 
To answer the research question, home ownership rates had to be calculated across generations. 
For this analysis, a home ownership rate was defined:

\[\text{Home Ownership Rate (%)} = 100 \cdot \frac{\text{Owner Occupied Households}}{\text{Total Occupied Households}}\]

\
This definition is consistent the definition given by the Census Housing Vacancy
Survey [2]. As the dataset provides weights for calculating household level
statistics, the home ownership rate was calculated as the the sum of the weights
associated with an owner occupied household divided by the sum of weights associated
with an occupied household. Finally, a generations home ownership rate is
calculated the same, except only households owned/occupied by an individual in
that generation are counted.

\
To find owner occupied households within the dataset both the OWNERSHP and RELATE variable had
to be used. The OWNERSHP variable in the dataset refers to whether the household the observation
belongs to is owned or rented; this does not mean that the individual is a 
homeowners as, for example, an individual living with parents who own a home would
have a value of ownership rather than rented in the OWNERSHIP variable. 
Instead, both the variables OWNERSHP and RELATE have to be used to find homeowners. 
If OWNERSHP indicates that the observation
belong to a household that owns their home and RELATE indicates that the observation
is the householder then that observation is considered a homeowner.

\
Finally, visualizations and summaries were created to explore the differences in 
homeownership rates between generations. 

\
First, table summaries were created
to show homeownership rates by generation in different years and at different ages. 
To create the table of homeownership rate by year between generations, homeownership
rates were calculated for each year grouped by generation. To create the table of
homeownership rate by age between generations, homeownership rates
were calculated for each age grouped by generation; for example, the homeownership
rate for 25 year old Baby Boomers would be calculated using only observations
were the age was 25 and the observation was a Baby Boomer (these would be from
years between 1971-1989).

\
To further explore the trend of homeownership rate over time, a line chart was created to show 
homeownership by year with a trend line for each generation. 
Similarly, to explore the trend of homeownership by age, a line chart was created 
to show homeownership by age with a trend line for each generation.

\
To explore difference in homeownership in locations by generation, a map visualization
was created to show home ownership at the state level, and at the county level. To
create these map visualizations, geojson files of USA state and county maps 
that included FIPS data were used. Home ownership rates were calculated by
state/county and then this data was merged into the geojson files using the
FIPS codes. For the county map, some counties were not identified, these
counties were merged together within state borders to form a map that 
contained homeownerhip rates in all identified counties, and home ownership
rates in states not including identified counties within the state. Similarly,
any county with <50 observations in any generation was combined with the
non-identified counties within state borders.

```{r warning=FALSE, echo=FALSE, output='none', message=FALSE}
library(ipumsr)
library(raster)
library(tidyverse)
library(gridExtra)
library(httr)
library(leaflet)
library(kableExtra)
```

```{r eval=FALSE, echo=FALSE}
IPUMS_API_KEY = "[API KEY]"
```

```{r echo=FALSE, eval=FALSE}
# create body of post request as JSON string
body <- '{
    "description": "API Extract",
    "data_structure": { 
        "rectangular": {
            "on": "P"
        }
    },
    "data_format": "fixed_width",
    "variables": {
      "YEAR":{},
      "AGE":{},
      "ASECWTH":{},
      "RELATE":{},
      "OWNERSHP":{},
      "STATEFIP": {},
      "COUNTY": {}
    },
    "samples": {'
# loop over years to add them all to "samples" part of body
for (year in 1976:2022) body <- paste0(body, "\"cps", year, "_03s\"", ":{},")
body <- paste0(body, "\"cps2023_03s\"", ":{}}}")

# post request with body
post_request <- POST(
  url = "https://api.ipums.org/",
  path = "extracts",
  query = list(
    "collection" = "cps",
    "version" = "beta"
  ),
  config = add_headers(
    "Authorization" = IPUMS_API_KEY,
    "Content-Type" = "application/json"
  ),
  body = body
)
```

```{r eval=FALSE, echo=FALSE}
# repeat checking if the request is complete
get_content <- NULL
time_between_check <- 60
repeat{
  # get request
  get_request <- GET(
    url = "https://api.ipums.org/",
    path = "extracts",
    config = add_headers(
      "Authorization" = IPUMS_API_KEY,
      "Content-Type" = "application/json"
    ),
    query = list(
      "collection" = "cps",
      "version" = "beta"
    )
  )
  
  # check if request is completed
  get_content <- content(get_request)[[1]] #get the most recent request content
  if (get_content$status == "completed") break
  # if not wait 30s and then try again
  Sys.sleep(time_between_check)
}

# get urls of xml and data
data_link <- get_content$download_links$data$url
xml_link <- get_content$download_links$ddi_codebook$url

# write .dat to directory
data_get <- GET(
  url = data_link,
  config = add_headers(Authorization=IPUMS_API_KEY),
  write_disk("data_file.dat.gz", overwrite=TRUE)
)

# write .xml to directory
xml_get <- GET(
  url = xml_link,
  config = add_headers(Authorization=IPUMS_API_KEY),
  write_disk("xml_file.xml", overwrite=TRUE)
)
```

```{r echo=FALSE, output='none', warning=FALSE, message=FALSE}
# Use IPUMS functions to get data table
ddi <- read_ipums_ddi("./xml_file.xml")
data <- read_ipums_micro(ddi, data_file="./data_file.dat.gz")
```

```{r eval=FALSE, echo=FALSE}
# check data dimension, head, tail, and variables
head(data)
tail(data)
dim(data)
str(data)
```

```{r eval=FALSE, echo=FALSE}
# key variables are AGE, RELATE, OWNERSHP, ASECWTH, YEAR
data <- data %>% select(AGE, RELATE, OWNERSHP, ASECWTH, YEAR, STATEFIP, COUNTY)
```

```{r eval=FALSE, echo=FALSE}
# key variables
summary(data$AGE)
summary(data$YEAR)
table(data$RELATE)
table(data$OWNERSHP)
table(data$STATEFIP)
table(data$COUNTY)
```

```{r eval=FALSE, echo=FALSE}
# further investigation of county NA (when do they occur?)
data %>%
  filter(is.na(COUNTY)) %>%
  group_by(YEAR) %>%
  summarise(count=n()) %>%
  ggplot(aes(x=YEAR, y=count)) +
  geom_bar(stat='identity')
```

```{r eval=FALSE, echo=FALSE}
# check distribution of age
data %>%
  ggplot(aes(x=AGE)) +
  geom_histogram()
```

```{r eval=FALSE, echo=FALSE}
# check old outliers
data %>%
  filter(AGE > 90)
```

```{r, echo=FALSE}
# filter younger than 18
data <- data %>% filter(AGE >= 19 & AGE <= 90)
```

```{r eval=FALSE, echo=FALSE}
# check number observations after removing <18
dim(data)
```

```{r eval=FALSE, echo=FALSE}
# what is the earliest generation in the dataset?

min(data$YEAR - data$AGE) #earliest birthday

data %>% # table of counts by birthyear
  mutate(BIRTHYEAR = YEAR-AGE) %>%
  group_by(BIRTHYEAR) %>%
  summarise(n())

data %>% # counts for gen before silent and all after
  mutate(BIRTHYEAR = YEAR-AGE) %>%
  filter(BIRTHYEAR < 1946) %>%
  summarise(n())
```

```{r echo=FALSE}
# create the generation variable
data <- data %>%
  mutate(GEN = ifelse( (YEAR-AGE >= 1946) & (YEAR-AGE <= 1964), "Baby Boomer",
               ifelse( (YEAR-AGE >= 1965) & (YEAR-AGE <= 1980), "Gen X",
               ifelse( (YEAR-AGE >= 1981) & (YEAR-AGE <= 1996), "Millenial",
               ifelse( (YEAR-AGE >= 1997) & (YEAR-AGE <= 2012), "Gen Z", 
               "Pre 1946")))))
```

```{r eval=FALSE, echo=FALSE}
# check counts by generation
data %>% 
  group_by(GEN) %>%
  summarise(n())
```

## Preliminary Results

```{r echo=FALSE}
data %>%
  filter(GEN != 'Pre 1946') %>%
  filter(YEAR %in% 2018:2023) %>%
  filter(RELATE == 0101) %>%
  group_by(YEAR, GEN) %>%
  reframe(RATE = 100*weighted.mean(OWNERSHP==10, ASECWTH)) %>%
  pivot_wider(names_from=GEN, values_from=RATE) %>%
  kable()
```

\
This table shows the homeownership rates by generation in recent years. Since
the older generations like Baby Boomers and Millennials have been alive longer,
and likely have more wealth, we would expect to see that their home ownership
rate is much higher as shown in the table. We also see that the home ownership
rate for Millennials has increased significantly in this time period, around 
a 14% difference with a signicant increase in 2020. Most Millennials would be
in their late 20s and early 30s which is around the age of the average first
time homebuyer in the United States. There were also historically low interest rates
between 2019-2021 which could have contributed to this increase. Finally, we
can also see that Baby Boomers were the only generation to have a decrease in
homeownership rates within this time period, this could be due to aging populations
selling their home to move in with family or into care homes. In the future, we
would expect to see the Baby Boomer home ownership rate continue to fall. Overall,
we see that Millennials and Gen Z have the largest increase in home ownership
rates within this time period.

```{r echo=FALSE}
data %>%
  filter(GEN != 'Pre 1946') %>%
  filter(AGE %in% seq(from=25, to=40, by=5)) %>%
  filter(RELATE == 0101) %>%
  group_by(AGE, GEN) %>%
  reframe(RATE = 100*weighted.mean(OWNERSHP==10, ASECWTH)) %>%
  pivot_wider(names_from=GEN, values_from=RATE) %>%
  kable()
```

\
This table shows the homeownership rates at certain ages by generation. As the
oldest members of Gen Z were only 26 in 2023, their is no data available in
the later ages. We can see that Baby Boomers have the highest home ownership
rate at all ages while Millennials have the lowest across all ages except
25. While Gen Z only has data at the age 25, it has the second highest home
ownership rate at this age. There does appear to be a trend of homeownership
rates dropping at all ages as the generations get younger; however, Gen Z could
disrupt this trend if their rates continue to be high. Without further data on
Gen Z, we cannot know if this trend will continue or Gen Z will outperform 
Millennials and/or Gen X.

```{r echo=FALSE}
# over all age groups
rate_by_time_all <- data %>%
  group_by(YEAR) %>%
  filter(RELATE == 0101) %>%
  summarise(RATE = 100*weighted.mean(OWNERSHP==10, ASECWTH)) %>%
  drop_na() %>%
  ggplot(aes(x=YEAR, y=RATE)) +
  geom_line(linewidth=1, color='black') +
  ylim(0, 100) +
  labs(y="Home Ownership Rate (%)", x="Year", title="Home Ownership Rate By Year")

rate_by_age_all <- data %>%
  group_by(AGE) %>%
  filter(RELATE == 0101) %>%
  summarise(RATE = 100*weighted.mean(OWNERSHP==10, ASECWTH)) %>%
  drop_na() %>%
  ggplot(aes(x=AGE, y=RATE)) +
  geom_line(linewidth=1, color='black') +
  ylim(0, 100) +
  labs(x="Age", title="Home Ownership Rate By Age") +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

grid.arrange(rate_by_time_all, rate_by_age_all, ncol=2)
```
\
These plots show overall trends of home ownership by year and age. We can see
that home ownership rates tend to stay fairly similar over time. This makes sense
as outside of large economic shifts we would expect to have the proportion of
people owning homes stay fairly constant as younger people buy houses and older
people move out.We can also see that home ownership rates by age increases over 
time and then starts to flatten out around 50, before starting to decrease at 
around 80. Again, this would make sense as people would want to buy homes once
they have a stable income and career which would likely occur in the age range
25-40. Then since most people who had the means to buy a house or wanted to buy
a house likely already did by 50, the rate starts to flatten. Finally, once
a person is older, they may start to sell their homes and move in with family
or care home. Interestingly, we also see an outlier at around 19 where the rate 
starts off higher. The slightly elevated rate at 19 could be due to inheritance where 
an individual is not legally allowed to own a property until they reach age of majority.

```{r echo=FALSE}
data %>%
  filter(GEN != 'Pre 1946') %>%
  filter(RELATE == 0101) %>%
  group_by(YEAR, GEN) %>%
  summarise(RATE = 100*weighted.mean(OWNERSHP==10, ASECWTH), .groups="rowwise") %>%
  drop_na() %>%
  ggplot(aes(x=YEAR, y=RATE, color=GEN)) +
  geom_line(linewidth=1) +
  ylim(0, 100) +
  theme(legend.position="bottom")
```
\
This plot shows the trend of home ownership by year between each generation.
We can see that Baby Boomer rate is has flattened significantly since
around 2005 and the Gen X was starting to flatten around 2005 as well
although it has started to increase again since around 2015; the sub prime mortgage
crisis occurred between 2007 and 2010 and could have contributed to this decrease
in the rate of change of the Gen X home ownership rate. For Millennials, we can 
see that the rate has been increasing and is starts to increase faster around
2015. Finally, for Gen Z there is not a lot of data, but we can see that it
is starting to increase and in 2020 it had the rate started to increase faster.

```{r echo=FALSE}
data %>%
  filter(GEN != 'Pre 1946') %>%
  filter(RELATE==0101) %>%
  group_by(AGE, GEN) %>%
  summarise(RATE = 100*weighted.mean(OWNERSHP==10, ASECWTH), .groups="rowwise") %>%
  drop_na() %>%
  ggplot(aes(x=AGE, y=RATE, color=GEN)) +
  geom_line(linewidth=1) +
  ylim(0, 100) +
  theme(legend.position="bottom")
```

\
This plot shows the trend of home ownership by age between each generation.
The plot provides a good comparison of home ownership rates by generation.
We can see that Baby Boomers have consistently outperformed all other generations
past the age of 26 and we can also see that Millenials have under preformed all
other generations past the age of 26. The Gen X rate tracks fairly closely to the
Baby Boomer rate before it starts to under preform in the 30s age range and onward.
Gen Z out preforms both Millennial and Gen X for the age range where data is available,
while the Gen Z line flattens towards the age of 26, without further data we do not
know whether Gen Z will outperform or under preform other generations. Low intrest
rates from 2019-2021, when older members of Gen Z would have started thinking about
buying a home, could have contributed to the high home ownership rates of 
Gen Z relative to the other generations.

```{r echo=FALSE}
# setup proportions of homeowners by state for different generations
states <- geojsonio::geojson_read("https://rstudio.github.io/leaflet/json/us-states.geojson", what = "sp")
map_states <- data.frame(id=as.numeric(states$id))
owners_by_state <- data %>% 
  filter(RELATE == 0101) %>%
  filter(GEN != 'Pre 1946') %>%
  group_by(GEN, STATEFIP) %>%
  reframe(p = round(100*weighted.mean(OWNERSHP==10, ASECWTH), digits=2)) %>%
  drop_na() %>%
  right_join(map_states, join_by(STATEFIP==id)) %>%
  pivot_wider(names_from=GEN, values_from=p) %>%
  select(!"NA")
states$boomer <- owners_by_state[match(map_states$id, owners_by_state$STATEFIP),]$"Baby Boomer"
states$millenial <- owners_by_state[match(map_states$id, owners_by_state$STATEFIP),]$"Millenial"
states$genx <- owners_by_state[match(map_states$id, owners_by_state$STATEFIP),]$"Gen X"
states$genz <- owners_by_state[match(map_states$id, owners_by_state$STATEFIP),]$"Gen Z"
```

```{r warning=FALSE, echo=FALSE, message=FALSE, output='none'}
bins <- c(0, 20, 30, 40, 50, 60, 70, 80, 90, 100)

# palettes
pal1 <- colorBin("YlOrRd", domain = states$boomer, bins = bins)
pal2 <- colorBin("YlOrRd", domain = states$genx, bins = bins)
pal3 <- colorBin("YlOrRd", domain = states$millenial, bins = bins)
pal4 <- colorBin("YlOrRd", domain = states$genz, bins = bins)

# generation labels
labelsBoomer <- sprintf(
  "<strong>%s</strong><br/>%g &#37 of Homeowners",
  states$name, states$boomer
) %>% lapply(htmltools::HTML)
labelsGenX <- sprintf(
  "<strong>%s</strong><br/>%g &#37 of Homeowners",
  states$name, states$genx
) %>% lapply(htmltools::HTML)
labelsMillenial <- sprintf(
  "<strong>%s</strong><br/>%g &#37 of Homeowners",
  states$name, states$millenial
) %>% lapply(htmltools::HTML)
labelsGenZ <- sprintf(
  "<strong>%s</strong><br/>%g &#37 of Homeowners",
  states$name, states$genz
) %>% lapply(htmltools::HTML)

# setup
state_leaflet <- leaflet(states) %>%
  setView(-96, 37.8, 4) %>%
  addProviderTiles("MapBox", options = providerTileOptions(
    id = "mapbox.light",
    accessToken = Sys.getenv('MAPBOX_ACCESS_TOKEN'))) 

# Baby Boomers
state_leaflet <- state_leaflet %>%
  addPolygons(
    group = "Baby Boomer",
    fillColor = ~pal1(boomer),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labelsBoomer,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"))

# Gen X
state_leaflet <- state_leaflet %>%
  addPolygons(
    group = "Gen X",
    fillColor = ~pal2(genx),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labelsGenX,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"))

# Millenial
state_leaflet <- state_leaflet %>%
  addPolygons(
    group = "Millenial",
    fillColor = ~pal3(millenial),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labelsMillenial,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"))

# Gen Z
state_leaflet <- state_leaflet %>%
  addPolygons(
    group = "Gen Z",
    fillColor = ~pal4(genz),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labelsGenZ,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"))

state_leaflet %>%
  addLegend(pal = pal1, values = ~boomer, opacity = 0.7, title = NULL,
    position = "bottomright") %>%
  addLayersControl(
    baseGroups = c("Baby Boomer", "Gen X", "Millenial", "Gen Z"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

\
This map shows home ownership rates by state and generation. We can see that 
Baby Boomers have fairly high home ownership rates across all states. The 
states with the lowest home ownership rates for Baby Boomers are California,
New York, and Hawaii. For Gen X, we see lower home ownership than Baby Boomers.
We also see that Midwestern states and non-coastal states tend to have higher
home ownership rates, which could be due to affordability. For Millennials, we 
can see again that home ownership is lower than Gen X, but we also see that 
the states with the lowest home ownership rates are similar to the states with 
the lowest home ownership rates for Gen Z. Finally, for Gen Z the states with
the lowest home ownership rates are fairly similar to Millenials and Gen X, 
but surprisingly Florida is one of the states with the highest home ownership
rate which is unique to Gen Z.

```{r echo=FALSE}
# setup proportions of homeowners by county for different generations
counties <- geojsonio::geojson_read("./counties.geojson", what = "sp")
map_codes <- data.frame(fip = as.numeric(paste0(counties$STATEFP, counties$COUNTYFP)))
owners_by_county <- data %>% 
  filter(GEN != 'Pre 1946') %>%
  filter(RELATE == 0101) %>%
  group_by(COUNTY, GEN) %>%
  reframe(p = round(100*weighted.mean(OWNERSHP==10, ASECWTH), digits=2)) %>%
  filter(COUNTY != 0 & !is.na(COUNTY)) %>%
  right_join(map_codes, join_by(COUNTY==fip)) %>%
  pivot_wider(names_from=GEN, values_from=p) %>%
  select(!"NA")
counties$boomer <- owners_by_county[match(map_codes$fip, owners_by_county$COUNTY),]$"Baby Boomer"
counties$millenial <- owners_by_county[match(map_codes$fip, owners_by_county$COUNTY),]$"Millenial"
counties$genx <- owners_by_county[match(map_codes$fip, owners_by_county$COUNTY),]$"Gen X"
counties$genz <- owners_by_county[match(map_codes$fip, owners_by_county$COUNTY),]$"Gen Z"
```

```{r echo=FALSE}
# counties < 100 observations in any generation
include_county <- data %>%
  filter(GEN != 'Pre 1946') %>%
  filter(!is.na(COUNTY) & COUNTY != 0) %>%
  group_by(COUNTY, GEN) %>%
  summarise(n = n(), .groups = 'rowwise') %>%
  pivot_wider(names_from=GEN, values_from = n) %>%
  drop_na() %>%
  filter(.[[1]] >= 50, .[[2]] >= 50, .[[3]] >= 50, .[[4]] >= 50) %>%
  select(COUNTY)
codes <- paste0(counties@data$STATEFP, counties@data$COUNTYFP)
```

```{r echo=FALSE}
max = as.numeric(counties@polygons[[length(counties@polygons)]]@ID)
spoly <- list()
ids <- list()
mapdata <- data.frame(matrix(1, ncol=ncol(counties@data), nrow=0))
colnames(mapdata) <- colnames(counties@data)

vec <- c('Baby Boomer', 'Millenial', 'Gen X', 'Gen Z')
fips <- data.frame(id=as.integer(states@data$id), name=states@data$name)
statedata <- data %>%
    filter(GEN != "Pre 1946") %>%
    filter(RELATE == 0101 & COUNTY==0) %>%
    group_by(GEN, STATEFIP) %>%
    reframe(x1=1, x2=1, x3=1, x4=1, temp=1, x5=1, x6=1, x7=1, 
            p = 100*round(weighted.mean(OWNERSHP==10, ASECWTH), digits=2)) %>%
    arrange(match(GEN, vec)) %>%
    pivot_wider(names_from = GEN, values_from = p) %>%
    right_join(fips, join_by(STATEFIP==id)) %>%
    mutate(temp = paste0(name, " w/o Counties")) %>%
    select(-name) %>%
    arrange(STATEFIP) %>%
    as.data.frame()

for (i in 1:51) {
  
  if (i==2) next #skip alaska since it crashes r
  
  # state polygon setup
  x <- states@polygons[[i]]
  pid <- x@ID
  fip <- as.numeric(states@data[pid,"id"])
  name <- states@data[pid,"name"]
  x@ID <- as.character(as.numeric(x@ID)+max)
  pid <- x@ID
  x <- SpatialPolygons(list(x))
  
  # county polygon setup
  y <- counties
  
  idx <- intersect(
    which(as.numeric(y$STATEFP)==fip),
    which(!(is.na(counties$boomer) | is.na(counties$genx) | is.na(counties$millenial) | is.na(counties$genz)))
  )
  
  idx <- intersect(
    which(as.numeric(y$STATEFP)==fip),
    which(str_remove(codes, "^0+") %in% include_county$COUNTY)
  )

  y <- SpatialPolygons(y@polygons[idx])
  
  #
  if (length(idx)==0) z <- x else  z <- x - y
  
  UNNESTED <- list()
  unnest_polygons <- function(a) {
    if (class(a)=="Polygons") {
      lapply(a@Polygons, unnest_polygons)
    } else if (class(a)=="Polygon") {
      UNNESTED <<- c(UNNESTED, a)
    }
  }
  
  lapply(z@polygons, unnest_polygons)
  z <- Polygons(UNNESTED, ID=pid)
  z <- list(z)
  
  mapdata[pid,] = statedata[statedata$STATEFIP==fip,]
  spoly <- c(spoly, z)
  ids <- c(ids, pid)
  
}
```

```{r echo=FALSE}
idx <- intersect(
  which(!(is.na(counties$boomer) | is.na(counties$genx) | is.na(counties$millenial) | is.na(counties$genz))),
  which(str_remove(codes, "^0+") %in% include_county$COUNTY)
)
idx <- which(str_remove(codes, "^0+") %in% include_county$COUNTY)

state_county <- SpatialPolygons(counties@polygons[idx])
state_county <- SpatialPolygonsDataFrame(state_county, counties@data[idx,])

state_county@polygons <- c(state_county@polygons, spoly)
state_county@data <- rbind(state_county@data, mapdata)
state_county@plotOrder <- c(as.integer(seq(from=length(state_county@plotOrder)+1, 
                                to=length(state_county@plotOrder)+length(spoly),
                                by=1)), state_county@plotOrder)
```

```{r warning=FALSE, echo=FALSE, message=FALSE}
bins <- c(0, 20, 30, 40, 50, 60, 70, 80, 90, 100)

# palettes
pal <- colorBin("YlOrRd", domain = state_county$boomer, bins = bins)

# generation labels
labelsBoomer <- sprintf(
  "<strong>%s</strong><br/>%g &#37 of Homeowners",
  state_county$NAME, state_county$boomer
) %>% lapply(htmltools::HTML)
labelsGenX <- sprintf(
  "<strong>%s</strong><br/>%g &#37 of Homeowners",
  state_county$NAME, state_county$genx
) %>% lapply(htmltools::HTML)
labelsMillenial <- sprintf(
  "<strong>%s</strong><br/>%g &#37 of Homeowners",
  state_county$NAME, state_county$millenial
) %>% lapply(htmltools::HTML)
labelsGenZ <- sprintf(
  "<strong>%s</strong><br/>%g &#37 of Homeowners",
  state_county$NAME, state_county$genz
) %>% lapply(htmltools::HTML)

# setup
state_leaflet <- leaflet(state_county) %>%
  setView(-96, 37.8, 4) %>%
  addProviderTiles("MapBox", options = providerTileOptions(
    id = "mapbox.light",
    accessToken = Sys.getenv('MAPBOX_ACCESS_TOKEN'))) 

# Baby Boomers
state_leaflet <- state_leaflet %>%
  addPolygons(
    group = "Baby Boomer",
    fillColor = ~pal(boomer),
    weight = 1,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 1,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labelsBoomer,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"))

# Gen X
state_leaflet <- state_leaflet %>%
  addPolygons(
    group = "Gen X",
    fillColor = ~pal(genx),
    weight = 1,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 1,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labelsGenX,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"))

# Millennial
state_leaflet <- state_leaflet %>%
  addPolygons(
    group = "Millennial",
    fillColor = ~pal(millenial),
    weight = 1,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 1,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labelsMillenial,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"))

# Gen Z
state_leaflet <- state_leaflet %>%
  addPolygons(
    group = "Gen Z",
    fillColor = ~pal(genz),
    weight = 1,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 1,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labelsGenZ,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"))

state_leaflet %>%
  addLegend(pal = pal, values = ~boomer, opacity = 0.7, title = NULL,
    position = "bottomright") %>%
  addLayersControl(
    baseGroups = c("Baby Boomer", "Gen X", "Millennial", "Gen Z"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

\
Similar to the previous map, this map shows home ownership rates by
geography and generation. This map includes counties that were identified
in the dataset, many counties were not identified in the dataset as their
population was below a threshold were there are concerns about deanonymizing
the data. All counties that were not identified in the dataset and any
county with less than 50 observations in any generation are combined 
into the state and labeled "State w/o Counties". Similar patterns can be seen
to the previous state map; however, the difference between urban and rural
home ownership rates are more obvious in this map, with rural areas tending to
have higher home ownership rates across all generations.

## Summary

\
The analysis compared home ownership rates across generations by investigating
trends in home ownership rate over time, age, and geography. 

\
The trend of home ownership rate over years between generations was similar,
with fast growth followed by a flattening of the curve; however, some
differences between home ownership rate over year trend could be seen, such 
as the increase in Gen X home ownership rate slowing down between the years
2005 and 2015, potentially due to housing market conditions in this period.
While some differences were present, the overall trends appeared similar, 
showing that the trend home ownership rate over time has not significantly 
changed between generations.

\
The trend of home ownership rate over age between generations was also similar.
Although some deviations could be seen in the trends between generations.
Millennials tended to have the lowest home ownership rates at all ages when
compared to the other generations, while Baby Boomers tended to have the highest
rates at all ages. Gen X tracked closely with Baby Boomer between the ages 
19-35 but started to under preform at later ages. Surprisingly, Gen Z was
had the highest home ownership rates in ages 19-23 and the second highest
in ages 23-26 with Baby Boomers as the highest in that age range. The trends
between each generation being similar and the high home ownership rates of Gen Z
indicate that home ownership has not increased or declined significantly between
generations despite Baby Boomers and Gen X tending to have higher rates.

\
The analysis of home ownership rates by state and by county showed that
home ownership rates tended to be similar across geography between the 
generations. Home ownership rates in less populous states and counties, such as
Midwestern states or non-identified counties, tended to be higher across all generations; 
although there were some interesting differences, such as Florida being one of the
states with the highest home ownership rate for Gen Z compared to all other 
generations were Florida was not one of the states with high home ownership rate.

\
Overall, the analysis showed that home ownership rates by year, age, and 
geography have remained fairly similar between generations. Although
the home ownership rates do not appear to have increased or declined significantly,
this does not necessarily mean the housing market remained the same. This analysis
does not take into account any income, mortgage rates, or pricing data, so it
cannot give an accurate picture of the housing market or housing affordability.
Further analysis that include more economic data would provide more cohesive
information about housing market between generations.

## Citations:

\
IPUMS:
Sarah Flood, Miriam King, Renae Rodgers, Steven Ruggles, J. Robert Warren,
Daniel Backman, Annie Chen, Grace Cooper, Stephanie Richards, Megan
Schouweiler, and Michael Westberry. IPUMS CPS: Version 11.0 [dataset].
Minneapolis, MN: IPUMS, 2023. https://doi.org/10.18128/D030.V11.0

\
Generation Ranges:

[1] Dimock, M. (2019, January 17). Defining generations: Where millennials end and generation Z begins. Pew Research Center. https://www.pewresearch.org/short-reads/2019/01/17/where-millennials-end-and-generation-z-begins/  

\
Home Ownership Definition:

[2] Definitions and Explanations. United States Census. https://www.census.gov/housing/hvs/definitions.pdf 

\
Home Ownership as a voting issue:

[3] O’Donnell, K. (2022, August 2). Young people are pissed off: Housing crush sours millennial voters. Politico. https://www.politico.com/news/2022/08/02/housing-millennials-biden-economy-00047704  


\
Geojson for states:

https://rstudio.github.io/leaflet/json/us-states.geojson

\
Geojson for counties:

https://gist.github.com/sdwfrost/d1c73f91dd9d175998ed166eb216994a